---
title: "Univariate distribution dynamics in R:"
subtitle: "Using the ggridges Package"
author: "Carlos Mendez"
output: 
  html_notebook:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 4
    number_sections: true
    code_folding: "show"
    theme: "cosmo"
    highlight: "monochrome"
    df_print: "kable"
  github_document: default
always_allow_html: true
---


<style>
h1.title {font-size: 18pt; color: DarkBlue;} 
body, h1, h2, h3, h4 {font-family: "Palatino", serif;}
body {font-size: 12pt;}
/* Headers */
h1,h2,h3,h4,h5,h6{font-size: 14pt; color: #00008B;}
body {color: #333333;}
a, a:hover {color: #8B3A62;}
pre {font-size: 12px;}
</style>




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ConvergenceClubs) # convergence clubs
library(ggridges)

# Change the presentation of decimal numbers to 4 and avoid scientific notation
options(prompt="R> ", digits=4, scipen=999)
```



Suggested Citation: 

> Mendez Carlos (2020). Univariate distribution dynamics in R: Using the ggridges package. R Studio/RPubs. Available at <https://rpubs.com/quarcs-lab/univariate-distribution-dynamics>

This work is licensed under the Creative Commons Attribution-Share Alike 4.0 International License. 


# Load the Data

Let us use a [dataset](https://rdrr.io/cran/ConvergenceClubs/man/GDP.html) containing the per-capita GDP of 152 Countries over the 1970-2003 period. The data is from the package [`ConvergenceClubs`](https://rdrr.io/cran/ConvergenceClubs/api/)

```{r}
data(GDP)
GDP
```

# Data transformations

## Pivot from wide to long

```{r}
dat_long <- GDP %>% 
  pivot_longer(
    cols = starts_with("Y"),
    names_prefix = "Y",
    names_to = "year",
    values_to = "GDPpc"
  )
dat_long
```


## Take the log of GPDpc

```{r}
dat_long <- dat_long %>% 
  mutate(
    ln_GPDpc = log(GDPpc)
  )
```

## Focus on the last the years

```{r}
dat_long <- dat_long %>% 
  filter(
    year >= 1998
  )
```



# Plot distribution dynamics


```{r}
dat_long %>% 
  ggplot(aes(x = ln_GPDpc, y = year)) +
  geom_density_ridges() +
  theme_ridges()
```


## Scale parameter

A setting of `scale=1` means the tallest density curve just touches the baseline of the next higher one.  

```{r}
dat_long %>% 
  ggplot(aes(x = ln_GPDpc, y = year)) +
  geom_density_ridges(scale = 1) + 
  theme_ridges()
```


Larger `scale` values create more overlap.


```{r}
dat_long %>% 
  ggplot(aes(x = ln_GPDpc, y = year)) +
  geom_density_ridges(scale = 10) + 
  theme_ridges()
```

## Fill colors

```{r}
dat_long %>% 
  ggplot(aes(x = ln_GPDpc, y = year, fill = stat(x))) +
  geom_density_ridges_gradient(scale = 10) +
  scale_fill_viridis_c(name = "Ln of GPDpc") +
  theme_ridges()
```

## Map the probabilities directly onto color.


```{r}
dat_long %>% 
  ggplot(aes(
    x = ln_GPDpc,
    y = year,
    fill =  0.5 - abs(0.5 - stat(ecdf))
            )
        ) +
stat_density_ridges(
  geom = "density_ridges_gradient",
  calc_ecdf = TRUE,
  scale = 10
                    ) +
  scale_fill_viridis_c(name = "Tail probability",
                       direction = -1) +
  theme_ridges()
```




## Indicate quantiles

### Median

```{r}
dat_long %>% 
  ggplot(aes(x = ln_GPDpc, y = year)) +
  stat_density_ridges(quantile_lines = TRUE, 
                      quantiles = 2,
                      scale = 10) +
  theme_ridges()
```


### Any quantile

```{r}
dat_long %>% 
  ggplot(aes(x = ln_GPDpc, y = year)) +
  stat_density_ridges(quantile_lines = TRUE, 
                      quantiles = c(0.025, 0.975),
                      scale = 10,
                      alpha = 0.7) +
  theme_ridges()
```

### Color by Quartile


```{r}
dat_long %>% 
  ggplot(aes(x = ln_GPDpc,
             y = year,
             fill = factor(stat(quantile))
             )
         ) +
  stat_density_ridges(
    geom = "density_ridges_gradient", 
    calc_ecdf = TRUE,
    quantiles = 4, 
    quantile_lines = TRUE,
    scale = 10
    ) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_ridges()
```


## Further modifications

```{r}
dat_long %>% 
  ggplot(aes(x = ln_GPDpc,
             y = year,
             fill = factor(stat(quantile))
             )
         ) +
  stat_density_ridges(
    geom = "density_ridges_gradient", 
    calc_ecdf = TRUE,
    quantiles = 4, 
    quantile_lines = TRUE,
    scale = 10
    ) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_ridges() +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  coord_cartesian(clip = "off") 
```


```{r}
dat_long %>% 
  ggplot(aes(x = ln_GPDpc,
             y = year,
             fill = factor(stat(quantile))
             )
         ) +
  stat_density_ridges(
    geom = "density_ridges_gradient", 
    calc_ecdf = TRUE,
    quantiles = 4, 
    quantile_lines = TRUE,
    scale = 10
    ) +
  scale_fill_viridis_d(name = "Quartiles") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  coord_cartesian(clip = "off") + 
  theme_ridges(grid = FALSE, center_axis_labels = TRUE)
```


# References

- <https://wilkelab.org/ggridges/>
- <https://wilkelab.org/ggridges/articles/introduction.html>

END
